---
title: "Coursera reproducable research - Wk2 assignment"
author: "Marco"  
date: "10-5-2020"  
output: html_document  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r load_libs,ECHO=FALSE,results='hide',include=FALSE,message=FALSE}
#start with getting the right libraries, dont want to echo this
library( dplyr )
library( ggplot2)
```

## Loading and preprocessing the data
All work is done in the current working directory  

```{r}
#get the data
dest = "./activity.zip"
if( !file.exists( dest))
{
  fileurl = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file( fileurl, destfile = dest)
  unzip( dest )
}
file = "./activity.csv"
df = read.csv( file=file)
head( df )
```
## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.
```{r}
#remove the NA's'
df_new <- na.omit( df )

#aggregate per day
days <-group_by( df_new, date ) 
days <- summarize( days, steps=sum(steps))
hist( days$steps,
      breaks=30,
      col="green",
      main = "Steps/Day")

# mean & median for the figures per day
oldmean <- mean( days$steps)
oldmedian <- median( days$steps)
paste0( "Mean is ", oldmean, " Median is ", oldmedian)
```
## What is the average daily activity pattern?  
And which 5 minute interval has the highest number?  

```{r}
#average daily pattern; days on y-axis, 5 minute pattern
#aggregate per interval
interval <-group_by( df_new, interval )
interval <- summarize( interval, steps=mean(steps))

ggplot( interval, type="l", 
        aes(interval, steps) ) +
        geom_line()
paste0( "Interval with the highest number: ", interval[ max <- which.max( interval$steps ), 1 ] )
paste0( "Steps for Interval with the highest number: ", interval[ max <- which.max( interval$steps ), 2 ] )
```

## Imputing missing values
How many do we have? 
And we fill them with the means per interval.  
Because on day level we dont always see values, on interval level we do.  
We do this by an extra function.

Further new mean and median values are asked.

```{r}
paste0( "Nbr missing values: ", nrow( df ) - nrow( df_new ) )

#therefore we create this funtion
replacewithmean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))

#and chain this into again df_new
df_new <- df %>% 
            group_by(interval) %>% 
            mutate(steps= replacewithmean(steps))

#again aggregate per day
#shorter; but have to assign column names
days <- aggregate( df_new$steps, by=list(df_new$date), sum)
names(days)[1]  <- "date"
names( days) [2] <- "steps"
# and create a new histogram
hist( days$steps,
      breaks=30,
      col="green",
      main = "Steps/Day")

# mean & median for the figures per day
newmean <- mean( days$steps)
newmedian <- median( days$steps)
paste0( "OLD: Mean is ", oldmean, " Median is ", oldmedian)
paste0( "NEW: Mean is ", newmean, " Median is ", newmedian)
```

## Finally Do we see activity pattern difference in the current dataset?
transform the dates to real dates and determine if they are weekend or not

```{r}
df_new$date <- as.Date( df_new$date)
df_new$day <- weekdays( df_new$date)
df_new$weekend <- ifelse( df_new$day=="zondag"|df_new$day=="zaterdag", "weekend", "werkdag")

#and aggregate the data
interval <-aggregate( df_new$steps, by=list(df_new$weekend, df_new$interval), sum)
names(interval)[1] <- "weekend"
names(interval)[2] <- "interval"
names(interval)[3] <- "steps"
head( interval )
ggplot( interval, aes(x=interval, y=steps, color=weekend)) +
  geom_line() +
  ggtitle( "Weekend vs workday comparison")
```